A White Noise Approach to Evolutionary Ecology

### 2020-04-18 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
split main text and SM

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued stochastic processes. This approach allows us to compute the full implications of demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a quantitative genetic model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests selection gradients can be either positively or negatively correlated with competition coefficients depending on the ratio of interspecific trait variation to intraspecific trait variation. Hence, this approach contributes to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of stochastic models that underlie rigorous investigations of the relationship between feedbacks of biological processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document: 
    extra_dependencies:
      - float
      - flafter
    fig_caption: yes
    fig_height: 3
    fig_width: 4
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

@@

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=(m(\nu,x)-\bar m(t)) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}
This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which, for sexually reproducing populations, has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539; and @Johnson2005], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance. To clarify the distinction between this model and previous models of phenotypic evolution we refer to the PDE (\ref{eq1}) from which (\ref{p_PDE}) was derived as the Deterministic Asexual Gaussian allelic model with Abundance dynamics (abbreviated DAGA). Later, we will extend this model to include the effects of demographic stochasticity, which we refer to as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA).

@@

In SM \S\ref{SDE_DERIV}, we extend these results to include the effects of demographic stochasticity. The key realization in this direction is that a rigorous derivation of a SPDE that naturally generalizes DAGA and accounts for ecological interactions has been provided by Li [-@zeng1998absolute]. To understand how this SPDE was derived, we provide an informal discussion of measure-valued branching processes (which treat populations as sets of discrete individuals) and their diffusion limits, the so-called superprocesses (which treat population size as a continuously varying number). Similar to the approach taken by Champagnat, Ferriére and Méléard [-@Champagnat2006], we begin with a BBM that models populations as discrete sets of reproducing individuals whose vital rates depend on their trait value as well as the state of the entire population. Taking a large population size limit and keeping our assumption of single dimensional traits, we employ a pair of classical results that show, under the appropriate rescaling in time, phenotypic space and population density, a sequence of rescaled BBM converges to a limiting process that can be characterized by a SPDE [@Mlard1993; @zeng1998absolute]. The limiting processes of rescaled BBM have been referred to as measure-valued Markov processes [@dawson1993measure] or superprocesses [@alisonetheridge2000]. Under the simplifying assumptions inherited from our treatment of deterministic dynamics and the additional assumption that the magnitude of demographic stochasticity is independent of trait values, we obtain as a special case a relatively simple expression for an SPDE that generalizes DAGA. The simplicity of our special case allows us to use properties of space-time white noise processes to derive a set of SDE that generalize equations (\ref{N_det}), (\ref{detmean}) and (\ref{vardyn_det}) to include the effects of demographic stochasticity. Classical expressions for the effects of random genetic drift on the evolution of mean traits are obtained as a special case.

@@

One can think of white noise as the static seen on old television sets or infinitely detailed random dust spread across both time and space. From a more mathematical, yet still informal perspective, white noise can be thought of as a stochastic process. That is, we can picture white noise as a collection of random variables indexed by time and possibly space. In relation to Brownian motion, denoted by $W$, white noise can be interpreted of as the derivative of Brownian motion with respect to time, denoted $\dot W$. Since Brownian motion can be thought to take infinitesimally small Gaussian distributed jumps at each time point, this leads to the conceptualization of white noise as a collection of Gaussian distributed random variables. Figure \ref{wn} illustrates realizations of this conceptualized white noise in one (left) and two (right) dimensions.

@@

However, it turns out that realizations of white noise do not exist as functions in the classical sense. Indeed, since Brownian motion is nowhere differentiable with respect to time, white noise cannot be formally understood as a time derivative. Thus our notation $\dot W$ is only meant to aid intuition and not be taken as formal. A formal understanding is possible by considering white noise as a _measure_-valued process [@Dawson1975; @Walsh] or as a _generalized_ process that acts on classically defined functions or stochastic processes to return either random variables or stochastic processes [@Krylov1981; @DaPrato2014]. Since a measure-valued process can be defined from a generalized process and a generalized process can be defined from a measure-valued process, the distinction between the two is more or less a matter of perspective. However, we find the perspective of white-noise as a generalized process to be a more efficient route for developing heuristics to help with some routine calculations involved with deriving SDE from SPDE. Hence, the notion of a generalized process provides the general idea implemented here. Although the treatments of SPDE provided by Krylov and Rozovskii [-@Krylov1981] and Da Prato and Zabczyk [-@DaPrato2014] extend the theory of SDE to formally treat SPDE in a general and elegant fashion, they require the navigation of an enourmous amount of technical definitions and detailed proofs. To extract some particularly useful results from this theory relevant to our goal of synthesizing the stochastic dynamics of biological populations, we provide a streamlined approach by capitolizing on the solid ground these authors have established. For instance, instead of rigorously proving properties of white-noise, we simply define them to be so, taking solice in the fact that the technical details have been worked out elsewhere. In SM \S\ref{DPZ14} we show how our informal treatment is related to the rigorous treatment provided by Da Prato and Zabczyk [-@DaPrato2014].

@@

We place quotations in the above expression to emphasize its informal nature and that it should not be confused with classical Riemann integration. Following this definition of white noise, we compute its value by sampling it using $\mathscr{N}_2$ processes. For example, integrating white noise over a region $D\times[0,t]$, with $t>0$ and $D$ a bounded subset of $\mathbb{R}$, is equivalent to evaluating $\mathbf W_t(I_{D\times [0,+\infty)})$ for the deterministic process

@@

I_{D\times [0,+\infty)}(x,t)=\left\{\begin{matrix}

@@

\|I_{D\times[0,+\infty))}\|_t^2=\mathbb{E}\left(\int_0^t\int_\mathbb{R}I_{D\times [0,+\infty)}^2(x,s)dxds\right)=t\int_Ddx=t|D|<+\infty, 

@@

where $|D|$ denotes the length of $D$, we have $I_{D\times [0,+\infty)}\in\mathscr{N}_2$. Thus, using equations (\ref{exp_WN}) and (\ref{cov_WN}) and adopting the informal notation introduced in equation (\ref{informal}), we can write the following

@@

To relate these formulae to the common notation used for SDE, we write

@@

In real populations individuals are born and potentially reproduce before they ultimately die. These three events provide the basic ingredients of a branching process. Mathematical investigations of such processes have a relatively deep history [@Kendall1966]. The most simple branching process, known as the Galton-Watson process, describes the number of individuals alive at a given time $t\geq0$ as a non-negative integer [@Kimmel2015]. Feller [-@feller1951] introduced a formal method to approximate branching processes with diffusion processes which are continuous in state (i.e., population size is approximated as a continuous quantity). Since diffusion processes possess greater analytical tractability than branching processes, Feller's method, known as the diffusion limit, has acquired immense popularity particularly in the field of mathematical population genetics [@Ewens2004]. For over the past half of a century a great deal of accomplishments have been achieved in formalizing the diffusion limits of branching processes that describe populations of individuals occuring in some continuous space [@Watanabe1968; @Dawson1975; @Dawson1978; @Perkins1992; @Mlard1993; @edwinarendperkins1995; @zeng1998absolute; @Bertoin2003; @Etheridge2008; @Barton2019]. This space can represent geographic space or, relevant to our context, phenotypic space. In the following subsection, we describe the BBM process, which is a particularly important branching process with spatial stucture. This process has been very useful in the study of SPDE due to its simplifying assumption that individuals do not interact. However, this assumption imposes an unfortunate restriction by precluding the modelling of ecological interactions. We therefore follow our discussion of BBM with a review of a few important results on spatially structured branching processes that account for interactions.

@@

  1) __Branching rate:__ In our formulation of BBM we assume Lifetimes of individuals are exponentially distributed with death rate $\lambda>0$ and reproduction occurs simultaneously with death. Biologically, this implies individuals are semelparous An alternative formulation treats birth and death events separately to model iteroparity. However, under the appropriate rescaling, both approaches converge to the same diffusion limit. We therefore choose the former approach for the sake of simplicity.

  2) __Reproductive law:__ When a birth event occurs a random (possibly zero) number of offspring are left. The distribution of offspring left is called the reproductive law or branching mechanism. We denote the mean and variance in reproductive output by $\mathscr{W}$ and $V$ respectively. The case of $\mathscr{W}=1$ is referred to as the critical condition. Under the critical condition the probability that extinction occurs in finite time is equal to one.

@@

Unfortunately, just as with cream cheese spread across too much toast, the blob perspective of the population may exhibit some complicated spatial discontinuities. However, for spatial dimension $d=1$, it turns out that $\mathscr{X}_t$ is absolutely continuous with respect to the Lebesgue measure for each $t\geq0$ [@Konno1988; @Reimers1989]. This means that we can write $\mathscr{X}_t(D)=\int_D \nu(x,t)dx$ for some density process $\nu(x,t)$. Setting $\lambda=1$ and $m=\ln\mathscr{W}$ this density process satisfies the SPDE
\begin{equation}\label{neutral_SPDE}
\frac{\partial}{\partial t}\nu(x,t)=m\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)+\sqrt{V\nu(x,t)}\dot W(x,t).
\end{equation}
Since $\nu(x,t)$ is not generally differentiable in $x$ or $t$, the derivatives in expression (\ref{neutral_SPDE}) are taken in the _weak_ sense [for examples, see @lawrenceevans2010]. That is, to rigorously interpret SPDE (\ref{neutral_SPDE}), we integrate the solution $\nu(x,t)$ against functions $f\in C_b^2(\mathbb{R})$, where $C_b^2(\mathbb{R})$ is the set of bounded and twice continuously differentiable functions on $\mathbb{R}$. Hence, equation (\ref{neutral_SPDE}) is just an abbreviation for

@@

A convergence theorem for the diffusion limit of a generalization of BBM was established by Watanabe [-@Watanabe1968]. Dawson [-@Dawson1975] suggested that, for spatial dimension $d=1$, this diffusion limit should admit a density process that satisfies a SPDE. Konna and Shiga [-@Konno1988] and Reimers [-@Reimers1989] independently proved Dawson's suggestion was indeed correct. The diffusion limit of this more general branching process (in arbitrary spatial dimension) is referred to as a Dawson-Watanabe superprocess [@alisonetheridge2000]. Conditioning a Dawson-Watanabe superprocess to have constant mass returns a Fleming-Viot process [@Etheridge1991; @Perkins1991] which has been popular in studies of spatial population genetics. In particular, an extension of the Fleming-Viot process, known as the $\Lambda$-Fleming-Viot process, was introduced by Bertoin and Le Gall [-@Bertoin2003] and coined by Etheridge [-@Etheridge2008] where it was used to resolve some technical challenges in modelling isolation by distance [@Felsenstein1975; see also @Barton2013; and @Barton2019]. Although this provides an impressive list of accomplishments, the Dawson-Watanabe superprocess falls short of our needs. In particular this process assumes individuals do not interact and thus precludes its ability to model nonlocal effects on the fitness of individuals, such as those produced via ecological interactions. However, this concern has been addressed, leading to constuctions of superprocesses that account for interactions among individuals. In the next subsection we summarize the main results in this area and introduce the SPDE that provides the basis for our approach to theoretical evolutionary ecology.

@@

Interactions that manifest in the spatial movement can be used to model mutation bias and those manisfesting in the reproductive law can model density-dependent growth rates and frequency-dependent selection. Perkins [-@Perkins1992; -@edwinarendperkins1995] developed a theory of stochastic integration with respect to the so-called _Brownian trees_ to characterize interacting superprocesses and establish properties of existence and uniqueness. Li [-@zeng1998absolute] built directly off of the construction of Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] to study properties of density processes associated with interacting superprocesses, arriving at a SPDE that forms the foundation of our approach.

Recall, we use $\nu(x,t)$ to denote the density of a superprocess, given it exists. Assuming the interactions manifest only in the reproductive law and that spatial movement follows Brownian motion with diffusion parameter $\sqrt\mu$ independent of both $X_t$ and $x$, Li [-@zeng1998absolute] proved a result that implies the interacting superprocess on one dimensional trait space has a density $\nu(x,t)$ which is non-negative, integrable, continuous in time and space and satisfies the SPDE
\begin{equation}\label{SPDE}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)+\sqrt{V\nu(x,t)}\dot W(x,t).
\end{equation}
As mentioned above, we refer to SPDE (\ref{SPDE}) as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA). Comparing equation (\ref{SPDE}) to equation (3.5) of Li [-@zeng1998absolute], our $m$ and $V$ correspond to Li's $b$ and $c$ respectively. It is important to note that, under the assumptions made in Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] and Li [-@zeng1998absolute], equation (\ref{SPDE}) is only formal when $m(h,x)$ is bounded across all combinations of $h\geq0$ and $x\in\mathbb{R}$. However, recalling our condition $m(h,x)\leq r\in\mathbb{R}$, the growth rates we consider are only bounded above. Yet, in the proof of the construction of the interacting superprocess as the limit of rescaled branching diffusions, Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] assumed $m(h,x)$ to be bounded to guarantee the total mass process will have finite mean and variance, for finite $t\geq0$. This allowed the authors to employ a tightness criterion for sequences of measures and show the rescaled processes converge to a superprocess with finite total mass. Li's [-@zeng1998absolute] result builds directly off of Méléard and Roelly's construction, inheriting the assumption of boundedness for $m(h,x)$. However, in Li [-@zeng1998absolute], the sufficiency of $m(h,x)$ being bounded above is even more clear since Li works explicitly with a common upperbound for both $m(h,x)$ and $V$. Hence, one can repeat the necessary proofs replacing the assumption that $m(h,x)$ is bounded with the assumption that $m(h,x)$ is merely bounded above to derive the same results. An alternative approach to establishing this result that allows for our relaxed assumption on the growth rate has carried out by Champagnat, Ferrière, and Méléard [-@Champagnat2006] using techniques developed by Evans and Perkins [-@Evans1994].

With solutions to SPDE (\ref{SPDE}) well defined for any growth rate bounded above, we can calculate the total mass process $N(t)$ using the mild solution of (\ref{SPDE}) with $f(x)\equiv1\in C_b^2(\mathbb{R})$ (the symbol "$\equiv$" means equal to for every $x$). That is,

>>>>

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued stochastic processes. This approach allows us to compute the full implications of demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a quantitative genetic model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests absolute values of linear selection gradients are negatively correlated with competition coefficients when the ratio of interspecific trait variation to intraspecific trait variation is approximately less than $1.25$ and positively correlated otherwise. The inverse relationship holds for quadratic selection gradients except with threshold diversity ratio equal to unity instead of $1.25$. Hence, this approach contributes to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of stochastic models that underlie rigorous investigations of the relationship between feedbacks of biological processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document: 
    extra_dependencies:
      - float
      - flafter
    fig_caption: yes
    fig_height: 3
    fig_width: 4
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{xr} # for cross-referencing between documents
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
  - \externaldocument{SM} # externel document to cross-reference
editor_options: 
  chunk_output_type: console
---

For cross-referencing to work we need the .aux files.
Hence, we need to run:

  tinytex::pdflatex("SM.tex", clean=FALSE)
  tinytex::pdflatex("draft.tex", clean=FALSE)
  tinytex::pdflatex("SM.tex", clean=FALSE)

@@

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=\Big(m(\nu,x)-\bar m(t)\Big) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}
This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which, for sexually reproducing populations, has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539; along with @Johnson2005], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance. To distinguish this model from previous models of phenotypic evolution we refer to PDE (\ref{eq1}) from which (\ref{p_PDE}) was derived as the Deterministic Asexual Gaussian allelic model with Abundance dynamics (abbreviated DAGA). Later, we will extend this model to include the effects of demographic stochasticity, which we refer to as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA).

@@

In SM \S\ref{SDE_DERIV}, we extend these results to include the effects of demographic stochasticity. The key realization in this direction is that a rigorous derivation of a SPDE that naturally generalizes DAGA and accounts for ecological interactions has been provided by Li [-@zeng1998absolute] and slightly generalized by Champagnat, Ferriére and Méléard [-@Champagnat2006]. To understand how this SPDE was derived, we provide an informal discussion of measure-valued branching processes (which treat populations as sets of discrete individuals) and their diffusion limits, the so-called superprocesses (which treat population size as a continuously varying number). Similar to the approach taken by Champagnat, Ferriére and Méléard [-@Champagnat2006], we begin with a BBM that models populations as discrete sets of reproducing individuals whose vital rates depend on their trait value as well as the state of the entire population. Taking a large population size limit and keeping our assumption of single dimensional traits, we employ a pair of classical results that show, under the appropriate rescaling in time, phenotypic space and population density, a sequence of rescaled BBM converges to a limiting process that can be characterized by a SPDE [@meleard1992interacting; @Mlard1993; @zeng1998absolute; @Champagnat2006]. The limiting processes of rescaled BBM have been referred to as measure-valued Markov processes [@dawson1993measure] or superprocesses [@alisonetheridge2000]. Under the simplifying assumptions inherited from our treatment of deterministic dynamics and the additional assumption that the magnitude of demographic stochasticity is independent of trait values, we obtain as a special case a relatively simple expression for an SPDE that generalizes DAGA. The simplicity of our special case allows us to use properties of space-time white noise processes to derive a set of SDE that generalize equations (\ref{N_det}), (\ref{detmean}) and (\ref{vardyn_det}) to include the effects of demographic stochasticity. Classical expressions for the effects of random genetic drift on the evolution of mean traits are obtained as a further special case.

@@

One can think of white noise as the static seen on old television sets or infinitely detailed random dust spread across both time and space. From a more mathematical, yet still informal perspective, white noise can be thought of as a stochastic process. That is, we can picture white noise as a collection of random variables indexed by time and possibly space. In relation to Brownian motion, denoted by $W$, white noise can be interpreted of as the derivative of Brownian motion with respect to time, denoted $\dot W$. Since Brownian motion can be thought to take infinitesimally small Gaussian distributed jumps at each time point, this leads to the conceptualization of white noise as a collection of independent Gaussian distributed random variables. Figure \ref{wn} illustrates realizations of this conceptualized white noise in one (left) and two (right) dimensions.

@@

However, it turns out that realizations of white noise do not exist as functions in the classical sense. Indeed, since Brownian motion is nowhere differentiable with respect to time, white noise cannot be formally understood as a time derivative. Thus our notation $\dot W$ is only meant to aid intuition and not be taken as formal. A formal understanding is possible by considering white noise as a _measure_-valued process [@Dawson1975; @Walsh] or as a _generalized_ process that acts on classically defined functions or stochastic processes to return either random variables or stochastic processes [@Krylov1981; @DaPrato2014]. Since a measure-valued process can be defined from a generalized process and a generalized process can be defined from a measure-valued process, the distinction between the two is more or less a matter of perspective. However, we find the perspective of white-noise as a generalized process to be a more efficient route for developing heuristics to help with some routine calculations involved with deriving SDE from SPDE. Hence, the notion of a generalized process provides the general idea implemented here. Although the treatments of SPDE provided by Krylov and Rozovskii [-@Krylov1981] and Da Prato and Zabczyk [-@DaPrato2014] extend the theory of SDE to formally treat SPDE in a general and elegant fashion, they require the navigation of an enourmous amount of technical definitions and detailed proofs. To extract some particularly useful results from this theory relevant to our goal of synthesizing the stochastic dynamics of biological populations, we provide a streamlined approach by capitolizing on the solid ground these authors have established. For instance, instead of rigorously proving properties of white noise, we simply define them to be so, taking solice in the fact that the technical details have been worked out elsewhere. In SM \S\ref{DPZ14} we show how our informal treatment is related to the rigorous treatment provided by Da Prato and Zabczyk [-@DaPrato2014].

@@

We place quotations in the above expression to emphasize its informal nature and that it should not be confused with classical Riemann integration. Following this definition of white noise, we compute its value by sampling it using $\mathscr{N}_2$ processes. For example, integrating white noise over a region $D\times[0,t]$, with $t>0$ and $D$ a bounded subset of $\mathbb{R}$, is equivalent to evaluating $\mathbf W_t(I_D)$ for the deterministic process

@@

I_D(x,t)=\left\{\begin{matrix}

@@

\|I_D\|_t^2=\mathbb{E}\left(\int_0^t\int_\mathbb{R}I_D^2(x,s)dxds\right)=t\int_Ddx=t|D|<+\infty, 

@@

where $|D|$ denotes the length of $D$, we have $I_D\in\mathscr{N}_2$. Thus, using equations (\ref{exp_WN}) and (\ref{cov_WN}) and adopting the informal notation introduced in equation (\ref{informal}), we can write the following

@@

To relate these formula to the common notation used for SDE, we write

@@

In real populations individuals are born and potentially reproduce before they ultimately die. These three events provide the basic ingredients of a branching process. Mathematical investigations of such processes have a relatively deep history [@Kendall1966]. The most simple branching process, known as the Galton-Watson process, describes the number of individuals alive at a given time $t\geq0$ as a non-negative integer [@Kimmel2015]. Feller [-@feller1951] introduced a formal method to approximate branching processes with diffusion processes which are continuous in state (i.e., population size is approximated as a continuous quantity). Since diffusion processes possess greater analytical tractability than branching processes, Feller's method, known as the diffusion limit, has acquired immense popularity particularly in the field of mathematical population genetics [@Ewens2004]. For over the past half of a century a great deal of accomplishments have been achieved in formalizing the diffusion limits of branching processes that describe populations of individuals occuring in some continuous space [@Watanabe1968; @Dawson1975; @Dawson1978; @Perkins1992; @Mlard1993; @edwinarendperkins1995; @zeng1998absolute; @Bertoin2003; @Etheridge2008; @Barton2019]. This space can represent geographic space or, relevant to our context, phenotypic space. In the following subsection, we describe the BBM process, which is a particularly important branching process with spatial structure. This process has been very useful in the study of SPDE due to its simplifying assumption that individuals do not interact. However, this assumption imposes an unfortunate restriction by precluding the modelling of ecological interactions. We therefore follow our discussion of BBM with a review of a few important results on spatially structured branching processes that account for interactions.

@@

  1) __Branching rate:__ In our formulation of BBM we assume Lifetimes of individuals are exponentially distributed with death rate $\lambda>0$ and reproduction occurs simultaneously with death. Biologically, this implies individuals are semelparous An alternative formulation treats birth and death events separately to model iteroparity. However, under the appropriate rescaling, both approaches converge to the same diffusion limit [@Champagnat2006]. We therefore choose the former approach for the sake of simplicity.

  2) __Reproductive law:__ When a birth event occurs we assume a random (possibly zero) number of offspring are produced. The distribution of offspring left is called the reproductive law or branching mechanism. We denote the mean and variance in reproductive output by $\mathscr{W}$ and $V$ respectively. The case of $\mathscr{W}=1$ is referred to as the critical condition. Since $X_t(\mathbb{R}^d)$ is a Galton-Watson branching process, it is well known the critical condition implies extinction in finite time with probability one, given finite $X_0(\mathbb{R}^d)$ [@Athreya1972].

@@

Unfortunately, just as with cream cheese spread across too much toast, the blob perspective of the population may exhibit some complicated spatial discontinuities. However, for spatial dimension $d=1$, it turns out that $\mathscr{X}_t$ is absolutely continuous with respect to the Lebesgue measure for each $t\geq0$ [@Konno1988; @Reimers1989]. This means that we can write $\mathscr{X}_t(D)=\int_D \nu(x,t)dx$ for some density process $\nu(x,t)$. Setting $\lambda=1$ and $m=\ln\mathscr{W}$ this density process satisfies the SPDE
\begin{equation}\label{neutral_SPDE}
\frac{\partial}{\partial t}\nu(x,t)=m\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)+\sqrt{V\nu(x,t)}\dot W(x,t).
\end{equation}
Since $\nu(x,t)$ is not generally differentiable in $x$ or $t$, the derivatives in expression (\ref{neutral_SPDE}) are taken in the _weak_ sense [for a definition see @lawrenceevans2010]. That is, to rigorously interpret SPDE (\ref{neutral_SPDE}), we integrate the solution $\nu(x,t)$ against functions $f\in C_b^2(\mathbb{R})$ where $C_b^2(\mathbb{R})$ is the set of bounded and twice continuously differentiable functions on $\mathbb{R}$. Hence, equation (\ref{neutral_SPDE}) is just an abbreviation for

@@

A convergence theorem for the diffusion limit of a generalization of BBM was established by Watanabe [-@Watanabe1968]. Dawson [-@Dawson1975] suggested that, for spatial dimension $d=1$, this diffusion limit should admit a density process that satisfies a SPDE. Konna and Shiga [-@Konno1988] and Reimers [-@Reimers1989] independently proved Dawson's suggestion was indeed correct. The diffusion limit of this more general branching process (in arbitrary spatial dimension) is referred to as a Dawson-Watanabe superprocess [@alisonetheridge2000]. Conditioning a Dawson-Watanabe superprocess to have constant mass returns a Fleming-Viot process [@Etheridge1991; @Perkins1991] which has been popular in studies of spatial population genetics. In particular, an extension of the Fleming-Viot process, known as the $\Lambda$-Fleming-Viot process, was introduced by Bertoin and Le Gall [-@Bertoin2003] and coined by Etheridge [-@Etheridge2008] where it was used to resolve some technical challenges in modelling isolation by distance [@Felsenstein1975; see also @Barton2013; and @Barton2019]. Although this provides an impressive list of accomplishments, the Dawson-Watanabe superprocess falls short of our needs. In particular this process assumes individuals do not interact and thus precludes its ability to model nonlocal effects on the fitness of individuals, such as those produced via competitive interactions. However, this concern has been addressed, leading to constuctions of superprocesses that account for interactions among individuals. In the next subsection we summarize the main results in this area and introduce the SPDE that provides the basis for our approach to theoretical evolutionary ecology.

@@

Interactions that manifest in the spatial movement can be used to model mutation bias and those manisfesting in the reproductive law can model density-dependent growth rates and frequency-dependent selection. Perkins [-@Perkins1992; -@edwinarendperkins1995] developed a theory of stochastic integration with respect to the so-called _Brownian trees_ to characterize interacting superprocesses and establish properties of existence and uniqueness. Li [-@zeng1998absolute] built directly off of the construction of Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] to study properties of interacting superprocesses and, by assuming individual spatial movement occurs independently of location $x$ and the entire population $X_t$, showed the evolution of associated density processes can be described by SPDE.

Recall $\nu(x,t)$ denotes the density of a superprocess, given it exists. Assuming the interactions manifest only in the reproductive law and that spatial movement follows Brownian motion with diffusion parameter $\sqrt\mu$ independent of both $X_t$ and $x$ and the growth rate is bounded both above and below, Li's [-@zeng1998absolute] result implies the interacting superprocess on one dimensional trait space has a density $\nu(x,t)$ which is non-negative, integrable, continuous in time and space and satisfies the SPDE
\begin{equation}\label{SPDE}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)+\sqrt{V\nu(x,t)}\dot W(x,t).
\end{equation}
As mentioned above, we refer to SPDE (\ref{SPDE}) as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA). Comparing equation (\ref{SPDE}) to equation (3.5) of Li [-@zeng1998absolute], our $m$ and $V$ correspond to Li's $b$ and $c$ respectively. Using techniques developed by Evans and Perkins [-@Evans1994], Champagnat, Ferrière and Méléard [-@Champagnat2006] generalized this result to the case when $m(h,x)$ is bounded above and is linear or more than linear in $h$. In particular, SAGA corresponds to a special case of equation (4.6) in Champagnat, Ferrière and Méléard [-@Champagnat2006] and is valid for the assumptions on $m$ made in \S\ref{deterministic}.

With solutions to SAGA well defined, we can calculate the total mass process $N(t)$ using the mild solution of (\ref{SPDE}) with $f(x)\equiv1\in C_b^2(\mathbb{R})$ (the symbol "$\equiv$" means equal to for every $x$). That is,
